Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-01 Paper/grass.tex
all

\begin{figure}
\includegraphics[scale=1]{Grass_Base_Nepf_shear}
\caption{Steady flow profile from the experiment of Ghisalberti\cite{Nepf04} (Case-I from Table-1 with 1250 plants/$m^2$, 
plant height = 13.7$\pm .2$cm and blade width of 0.64 cm)
 and it's approximation with $U_0=7.28$ cm/s and $\delta = 5.02cm$ with our model. The canopy extends upto $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the grass bed, and (iii) a boundary layer of thicness $\delta$ near the canopy top. The depedence of the boundary layer on the vegetation density $\Ndg$ is shown in the inset.}
\label{fig:basicflow}
\end{figure}

@@

In this paper we present a mathematical\shreyas model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear instability of a uniform flow over the grass bed. 
The threshold condition for the onset of the instability and the resulting waving frequency compares well with experimental observations.
In the limit of a dense grass bed, two distinct unstable modes can be asymptotically distinguished, with one of them sharing many characteristics with the Kelvin-Helmholtz instability. 
In the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, is considered to merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrate that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, in the interest of deriving the simplest mathematical model for the linear instability, we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, the flow instability is similar to the instability of a shear layer.

\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density
 parametrized by the boundary layer thickness. Parameters estimated from experiments reported by Ghisalberti and Nepf\cite{Ghisal02} to exhibit or suppress synchronous waving are 
also included in the top panel. In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed. Inset compares experimental observations 
of the dominant frequency $f_o$ (in Hz) measured with the predictions from the solution of \eqref{Orr-somerfield}. The experimental data in the inset in obtained from the articles by 
Ghisalberti \cite{Ghisal02} and Vivoni \cite{Vivoni98}.}
\label{Re_vs_delta}
\end{figure}
We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as $\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ \shreyas{citation needed} where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in figure \ref{Grass_Base_Vivoni_shear}, gives us a way to systematically control it and investigate its effect on the instability mechanism.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.68]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.68]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens46_imgsc}} \\
{\includegraphics[scale = 0.68]{Set6_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens34_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens36_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified grass number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

@@

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in Figure \ref{Re_vs_delta}. The observed frequencies are associated with the peaks in velocity spectra, frequency of monami and frequency of vortex passage \shreyas{cite someone}. The deviation of predicted frequency from the observed may be attributed to the various simplification we have made in our model. In a real canopy, the drag coefficients are known to
vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. The turbulent viscosity is also known to vary between the flow of unvegetated region and flow through the canopy. Although these variation
are not central to the mechanism leading to the instability of flow, we believe an improved analysis of flow with these variation in the model might lead to a better agreement between the
observed frequency and the predicted frequency.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by asymptotic analysis of \eqref{Orr-somerfield}.
Noting that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
Using this estimate, and $\Rey\gg1$, \eqref{Orr-somerfield} within the grass simplifies into $\sigma\left(D^2-k^2\right)\phi = 2{(\Ndg/\Rey)^{1/2}}D^2\phi$ \shreyas{check sign}, while in the unvegetated region it simplifies to the Rayleigh equation $ \left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$. 
This analysis predicts that the mode shape would converge as $\Rey\gg 1$ and $\Ndg \gg 1$ but $\Rey/\Ndg \sim O(1)$, in agreement with our numerical results shown in figure \ref{Asymptotic_mode}. 
We interpret this as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specifying in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 2 shapes appear self-similar in shape as $\delta\to 0$, and are compared to each other on a rescaled coordinate in the inset. Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. The limit is not yet reached for the case $\delta/H = 0.01$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}

>>>>

\begin{figure}
\includegraphics[scale=1]{Grass_Base_Nepf_shear}
\caption{Steady flow profile from the experiment of Ghisalberti\cite{Nepf04} (Case-I from Table-1 with 1250 plants/$m^2$, 
plant height = 13.7$\pm .2$cm and blade width of 0.64 cm)
 and it's approximation with $U_0=7.28$ cm/s and $\delta = 5.02cm$ with our model. The canopy extends upto $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the grass bed, and (iii) a boundary layer of thicness $\delta$ near the canopy top. 
The depedence of the boundary layer on the vegetation density $\Ndg$ is shown in the inset.}
\label{basicflow}
\end{figure}

@@

In this paper we present a mathematical \shreyas model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear instability of a uniform flow over the grass bed. 
The threshold condition for the onset of the instability and the resulting waving frequency compares well with experimental observations.
In the limit of a dense grass bed, two distinct unstable modes can be asymptotically distinguished, with one of them sharing many characteristics with the Kelvin-Helmholtz instability. 
In the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, is considered to merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, in the interest of deriving the simplest mathematical model for the linear instability, we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow (top) and the corresponding marginally stable wave number (bottom) for different submergence ration as a function of vegetation density
 parametrized by the boundary layer thickness. Parameters estimated from experiments reported by Ghisalberti and Nepf\cite{Ghisal02} to exhibit or suppress synchronous waving are 
also included in the top panel. In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed. Inset compares experimental observations 
of the dominant frequency $f_o$ (in Hz) measured with the predictions from the solution of \eqref{Orr-somerfield}. The experimental data in the inset in obtained from the aricles by 
Ghisalberti \cite{Ghisal02} and Vivoni \cite{Vivoni98}.}
\label{Re_vs_delta}
\end{figure}
We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as 
$\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ \shreyas{citation needed} 
where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{fig:basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.68]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.68]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.68]{Set5_dens46_imgsc}} \\
{\includegraphics[scale = 0.68]{Set6_dens32_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens34_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens36_imgsc}} &
{\includegraphics[scale = 0.68]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a given grass number density along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. For $\Ndg$ below
a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

@@

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in 
Figure \ref{Re_vs_delta}. The observed frequencies are associated with the peaks in velocity spectra, frequency of monami and frequency of vortex passage \cite{Ghisal02}\shreyas{cite someone}. The deviation of predicted frequency from the observed may be attributed to the various 
simplification we have made in our model. In a real canopy, the drag coefficients are known to
vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. The turbulent viscosity is also known to vary between the flow of unvegetated region and flow through the canopy\cite{Ghisal02}. Although these variation
are not central to the mechanism leading to the instability of flow, we believe an improved analysis of flow with these variation in the model might lead to a better agreement between the
observed frequency and the predicted frequency.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by asymptotic analysis of \eqref{Orr-somerfield}.
Noting that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
Using this estimate, and $\Rey\gg1$, \eqref{Orr-somerfield} within the grass simplifies into $\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi$ \shreyas{check sign}, while in the unvegetated region it simplifies to the Rayleigh equation $ \left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$. 
This analysis predicts that the mode shape would converge as $\Rey\gg 1$ and $\Ndg \gg 1$ but $\Rey/\Ndg \sim O(1)$, in agreement with out numerical results shown in figure \ref{Asymptotic_mode}. 
We interpret this as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{plot of $|\phi|$ for different shear layer thickness in limit of small shear layer thickness}
\label{Asymptotic_mode}
\end{figure}
